GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-10 manuscript.tex
outline comments
Update on Overleaf.

\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage[hidelinks]{hyperref,xcolor}
\renewcommand\UrlFont{\color{blue}\rmfamily}

@@

\author{
    Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} \and \\
    Third Author\inst{3}\orcidID{2222-3333-4444-5555}
}

@@

\begin{abstract}
The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
Web spatial data.\\

@@

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
possible handling of new spatial referencing systems, new geometry representations and new artifact 
presentation.\\

@@

>>>>

% Outline
% 
% Section 1
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile preentation
%       new URI regimes etc
% Section 5
%       expected changed use modes
\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage{todonotes}
\usepackage{listingsutf8}
\usepackage{cleveref}
\usepackage[hidelinks]{hyperref,xcolor}
\renewcommand\UrlFont{\color{blue}\rmfamily}

@@

\author{
    Nicholas J. Car\inst{1,2}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{3}\orcidID{0000-0002-9499-5840} % \and \\
    % Third Author\inst{3}\orcidID{2222-3333-4444-5555}
}

@@

\begin{abstract}
The Open Geospatial Consortium published the GeoSPARQL 1.0 standard \cite{open2012ogc} in 2012 containing multiple 
parts that define ``SPARQL extension functions'', ``RIF rules''\cite{kifer2013rif}, ``an RDF/OWL ontology for 
information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
Web spatial data.\\

@@

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter\cite{abhayaratna2020ogc} to: 
handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
possible handling of new spatial referencing systems, new geometry representations, and new artifact 
presentation.\\

@@

\section{Introduction}
\label{sec:introduction}
\section{Motivation to Update GeoSPARQL}
\label{sec:motivation}
Spatial Data On The Web Best Practices \cite{van2019best}
GeoSPARQL White Paper \cite{geosparqlwhitepaper}
\href{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}{https://www.w3.org/2015/spatial/wiki/Further\_development\_of\_GeoSPARQL}
\section{New features in GeoSPARQL 1.1}
\label{sec:newfeatures}
GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.
\subsection{New geometry literals}
\label{sec:literals}
GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} literal, and a DGGS\cite{sahr1998discrete} WKT Literal.
\small
\begin{lstlisting}[caption=GeoJSON literal example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
"{\"type\":\"Point\", \"coordinates\":[-83.38,33.95]}"
^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
\end{lstlisting}
\normalsize
Since the GeoJSON and KML formats are restricted to be represented in the WGS84 coordinate system only, these restrictions also apply to GeoJSON and KML literals. The DGGS WKT literal  
\subsection{Spatial aggregate functions}
\label{sec:spatialaggregate}
GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible outside of the semantic database, the inclusion of the functions provides distinct advantages:
\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}
GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.
Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum coordinates of a geometry respectively.
\subsection{Extensions of the GeoSPARQL ontology}
\label{sec:ontologyextensions}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, \href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength},
\href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the \href{http://www.opengis.net/ont/geosparql#inCRS}{geo:inCRS} property, which allows querying the coordinate reference system of a given geometry without the need to analyze geometry literal serializations. The property allows for the definition of a CRS as a URI. It paves the way for a future definition of coordinate reference systems fully in RDF as anticipated for GeoSPARQL 2.0.
\todo[inline]{I created a graphic (geold\_ontology.graphml) containing geo:SpatialObject, geo:Feature, geo:SpatialMeasure using the yEd Editor (https://www.yworks.com/products/yed) . If you find it useful we can extend it. }
\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{images/geold_ontology.png}
    \caption{GeoSPARQL 1.1 ontology including one example feature}
    \label{fig:geosparql11ontology}
\end{figure}
\section{Modernizing the documentation of GeoSPARQL}
\label{sec:documentation}
\subsection{Profile}
\label{sec:profile}
\section{Conclusions}
\label{sec:conclusions}
\subsection{Future Work}
\label{sec:futurework}
GeoSPARQL 1.2, GeoSPARQL 2.0?
\begin{itemize}
    \item GeoSPARQL extension ontologies: CRS systems
    \item GeoSPARQL 1.2: More literals?
    \item GeoSPARQL 2.0: Full featured support for 3D, simple features functions, coverages?
\end{itemize}
%   History of GeoSPARWL 1.0
% Section 2
%   motivation to update the spec
% Section 3
%   GeoSPARQL 1.1 additions
%       motivation, and expected use, of new spatial functions. In particular, why in GeoSPARQL when aggregations etc. can be performed elsewhere?
%       Spatial Measure
% Section 4
%       new profile preentation
%       new URI regimes etc
% Section 5
%       expected changed use modes
